Parsing Postfix log files

### 2009-03-20 doc/logparser-implementation.tex
Dealt with more XXX items.

Adding new actions is not as easy as adding new rules, though care has been
taken in the architecture and implementation to make adding new actions as
painless as possible.  The implementer writes a subroutine that accepts the
standard arguments given to actions, and registers it with the framework.
The new action must be registered before the rules are loaded, because it
is an error for a rule to specify an unregistered action; this helps catch
mistakes made when adding new rules.  No other work is required from the
implementer to integrate the action into the parser; all of their attention
and effort can be focused on correctly implementing their new action.  The
action may need to extend the list of valid combinations described in
\sectionref{out of order log lines} if the new action creates a different
set of acceptable programs, but this would only be necessary if the new
action processes log lines from Postfix components \parsername{} currently
does not have rules for.

@@

    \item [\_\_RESTRICTION\_START\_\_]  Matches the start of a delivery
        rejected log line: XXX SHOULD I INCLUDE THIS\@? \newline{}
        \verb!/(__QUEUEID__): reject(?:_warning)?: ! \newline{}
        \verb!(?:RCPT|DATA) from ! \newline{}
        \verb!(?>(__CLIENT_HOSTNAME__)\\[)! \newline{}
        \verb!(?>(__CLIENT_IP__)\\]): (__SMTP_CODE__)! \newline{}
        \verb!(?: __ENHANCED_STATUS_CODE__)?/!

@@

XXX IMPROVE THE INTRO\@: ``needs more of a lead-in''.

The complications described in this section are listed in the order in
which they were encountered during development of the parser.  Each of
these complications caused the parser to operate incorrectly, generating
either warning messages or leaving mails in the state table.  The frequency
of occurrence is much higher at the start of the list, with the first
complication occurring several orders of magnitude more frequently than the
last.  When deciding which problem to address next, the most common was
always chosen, as resolving the most common problem would yield the biggest
improvement in the parser, prune the greatest number of mails from the
state tables and error messages, and make the remaining problems more
apparent.  The first three complications were encountered early in the
parser's implementation and guided its design and development.

@@

The most difficult complication initially encountered is that locally
addressed mails are not always delivered directly to a mailbox: sometimes
they are addressed to and accepted for a local address but need to be
delivered to one or more remote addresses due to aliases.  When this
occurs a child mail will be injected into the Postfix queue, but without
the explicit logging that \daemon{smtpd} or \daemon{postdrop} injected
mails have.  Thus the source of the mail is not immediately discernible
from the log line in which the mail first appears; from a strictly
chronological reading of the log files it usually appears as if the child
mail has appeared from thin air.  Subsequently the parent mail will log the
creation of the child mail, e.g.\ parent mail 3FF7C4317 creates child mail
56F5B43FD\@:

\texttt{3FF7C4317: to=<username@example.com>, relay=local, \newline{}
\tab{} delay=0, status=sent (forwarded as 56F5B43FD)}

Unfortunately, while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition, the first log line belonging to the child mail (the log line
cited above belongs to the parent mail) is logged by \daemon{qmgr}, so the
order also depends on how soon \daemon{qmgr} processes the new mail.

Because of the uncertain order the parser cannot complain when it
encounters a log line from \daemon{qmgr} for a previously unseen mail;
instead it must flag the mail as coming from an unknown origin, and
subsequently clear the flag if and when the origin of the mail becomes
clear.  Obviously the parser could omit checking where mails originate
from, but requiring an explicit source helps to expose bugs in the parser;
such checks helped to identify the complications described in
\sectionref{discarding cleanup log lines} and \sectionref{pickup logging
after cleanup}.

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered, and entirely finished with
\textbf{before} the parent logs the creation log line!  Thus, mails flagged
as coming from an unknown origin cannot be entered into the database when
their final log line is parsed; instead they must be marked as ready for
entry and subsequently entered once the parent mail has been identified.

XXX MERGE THE REMAINDER OF THIS SECTION INTO THE PRIOR MATERIAL\@.

The crux of this complication is that re-injected mails appear in the log
files without explicit logging indicating their source.  There are two
implicit indications:

@@

    \item The indicator which more commonly introduces re-injection is when
        \daemon{qmgr} selects a mail with a previously unseen queueid for
        delivery, in which case a new data structure will be created for
        that mail.  The mail will be flagged as having unknown origins;
        this flag will subsequently be cleared once the origin has been
        established.  This may also be an indicator that the mail is a
        bounce notification, see \sectionref{identifying bounce
        notifications} for details.

    \item Local delivery re-injects the mail and logs a relayed delivery
        rather than delivering directly to a mailbox or program as it
        usually would.\footnote{Relayed delivery is performed by the
        \acronym{SMTP} client; local delivery means local to the server,
        i.e.\ an address the server is final destination for.}  In this
        case the mail may already have been created (described above) and
        the unknown origin flag will be cleared; if not a new data
        structure will be created.  In both cases the re-injected mail is
        marked as a child of the original mail.  The log line in question
        is:

        \texttt{3FF7C4317: to=<username@example.com>, relay=local,
        \newline{} \tab{} delay=0, status=sent (forwarded as 56F5B43FD)}

        This always occurs for re-injected mail but typically occurs after
        the first indicator.  This log line is required to tie the parent
        and child mails together and so is central to the process of
        tracking re-injected mails.

@@

The algorithm for tracking and saving re-injected mail to the database can
finally be described:

    \item If the mail is of unknown origin it is assumed to be a child mail
        whose parent has not yet been identified.  Mark the mail as ready
        for entry in the database and wait for the parent to deal with it.
        The mail should not have any subsequent log lines; only its parent
        should refer to it.

    \item If the mail is a child mail then it has already been tracked: as
        with all other mail, the data is cleaned up, the child is entered
        in the database, and then deleted from the state tables.  The child
        mail will be removed from the parent mail's list of children; if
        this is the last child and the parent has already been entered in
        the database, the parent will also be deleted from the state
        tables.

    \item The last alternative is that the mail is a parent mail.
        Regardless of the state of its children its data is cleaned up and
        entered in the database.  The parent may have children that are
        waiting to be entered in the database; the data for each of those
        children is cleaned up and entered in the database, then deleted
        from the state tables.  The parent may also have outstanding
        children which are not yet delivered, in which case the parent must
        be retained in the state tables until those children are finished
        with.  As soon as the last child is deleted from the state tables
        the parent will also be deleted from the state tables.

A parent mail can have multiple children, which may be delivered before or
after the parent mail.

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes there is a timeout or the
connection is lost\footnote{For brevity's sake \textit{timeout\/} will be
used throughout this section, but everything applies equally to lost
connections.} during the DATA phase; when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded with the connection data instead, which is saved.

To deal properly with timeouts the parsing algorithm needs to do the
following in the TIMEOUT action:

@@

    \item If one or more recipients have been accepted Postfix will have
        allocated a queueid for the incoming mail, and there will be a mail
        in the state tables that needs to be dealt with.

@@

XXX MAKE THIS PARAGRAPH CLEARER\@.  A timeout may thus apply either to an
accepted mail or a rejected mail.  To distinguish between the two cases the
algorithm compares the timestamp of the last accepted mail against the
timestamp of the last line logged by \daemon{smtpd} for that connection
(the TIMEOUT action is dependant on the CLONE action keeping a list of all
mails accepted on each connection).  If the mail acceptance timestamp is
later then the timeout applies to the just-accepted mail, which will be
discarded.  If the \daemon{smtpd} timestamp is later there was a rejection
between the accepted mail and the timeout: the action assumes that the
timeout applies to a rejected delivery attempt and finishes.  This
assumption is not necessarily correct, because Postfix may have accepted an
earlier recipient and rejected a later one, in which case the timeout
applies to the accepted mail, which should be discarded.  This has not been
a problem in practice, though it may be in future.  This complication is
further complicated by the presence of out of order \daemon{cleanup} log
lines: see \sectionref{discarding cleanup log lines} for details.

This complication is dealt with in the TIMEOUT action.

>>>>

Adding new actions is not as easy as adding new rules, though care has been
taken in the architecture and implementation to make adding new actions as
painless as possible.  The implementer writes a subroutine that accepts the
standard arguments given to actions, and registers it with the framework.
The new action must be registered before the rules are loaded, because it
is an error for a rule to specify an unregistered action; this helps catch
mistakes made when adding new rules.  No other work is required from the
implementer to integrate the action into the parser; all of their attention
and effort can be focused on correctly implementing their new action.  The
action may need to extend the list of valid combinations described in
\sectionref{out of order log lines} if the new action creates a different
set of acceptable programs, but this would only be necessary if the new
action processes log lines from Postfix components \parsername{} currently
does not have rules for, e.g.\ \daemon{virtual} or \daemon{lmtp}.

@@

    \item [\_\_RESTRICTION\_START\_\_]  Matches the standard information
        Postfix includes at the start of almost all log lines resulting
        from rejecting a delivery attempt.

@@

It was initially expected that parsing Postfix log files would be a
relatively simple task, requiring a couple of months of work.  The author
had found Postfix log files useful when investigating problems reported by
users, and an examination of the log files gave the impression that they
would be straightforward to parse, process, and understand.  The large
variation in log lines was not apparent, because the majority of log lines
are recognised by a small set of rules, as shown in \figureref{rule hits
graph}.  Most of the myriad complications and difficulties documented in
this section were discovered during \parsernames{} development, but the
first three complications were identified during the planning and design
phase of this project, influencing the architecture's design.

Each of these complications caused \parsername{} to operate incorrectly,
generate warning messages, or leave mails in the state table.  The
complications are listed in the order in which they were overcome during
development of \parsername{}, with the first complication occurring several
orders of magnitude more frequently than the last.  When deciding which
problem to tackle next, the problem causing the highest number of warning
messages or mails incorrectly remaining in the state tables was always
chosen, because that approach yielded the biggest improvement in the
parser, and made the remaining problems more apparent.

@@

Mails sent to local addresses are not always delivered directly to a
mailbox: sometimes they are sent to and accepted for a local address, but
need to be delivered to one or more remote addresses due to aliasing.  When
this occurs, a child mail will be injected into the Postfix queue, but
without the explicit logging that mails injected by \daemon{smtpd} or
\daemon{postdrop} have.  Thus the source of the mail is not immediately
discernible from the log line in which the mail's queueid first appears:
from a strictly chronological reading of the log lines it usually appears
as if the child mail has been created without a source.  Subsequently the
parent mail will log the creation of the child mail, e.g.\ parent mail
3FF7C4317 creates child mail 56F5B43FD\@:

\texttt{3FF7C4317: to=<username@example.com>, relay=local, \hfill{}
\newline{} \tab{} \tab{} delay=0, status=sent (forwarded as 56F5B43FD)}

Unfortunately, while all log lines from an individual process appear in
chronological order, the order in which log lines from different processes
are interleaved is subject to the vagaries of process scheduling.  In
addition, the first log line belonging to the child mail (the example log
line above belongs to the parent mail) is logged by \daemon{qmgr} or
\daemon{cleanup}, so the order also depends on how soon they process the
new mail.

Because of the uncertain order that log lines can appear in, \parsername{}
cannot complain when it encounters a log line from either \daemon{qmgr} or
\daemon{cleanup} for a previously unseen mail; instead it must flag the
mail as coming from an unknown origin, and subsequently clear that flag if
and when the origin of the mail becomes clear.  Obviously the parser could
omit checking where mails originate from, but requiring an explicit source
helps to expose bugs in the parser; such checks helped to identify the
complications described in \sectionref{discarding cleanup log lines} and
\sectionref{pickup logging after cleanup}.

Process scheduling can have a still more confusing effect: quite often the
child mail will be created, delivered, and entirely finished with,
\textit{before\/} the parent mail logs the creation log line!  Thus, mails
flagged as coming from an unknown origin cannot be entered into the
database when their final log line is processed; instead they must be
marked as ready for entry and subsequently entered once their origin has
been identified.  The crux of this complication is that re-injected mails
appear in the log files without explicit logging indicating their source.

Tracking re-injected mail requires \parsername{} to do the following in the
COMMIT action:

@@

    \item If a mail is tagged with the unknown origin flag it is assumed to
        be a child mail whose parent has not yet been identified.  The mail
        is tagged as ready to be entered in the database, but entry is
        postponed until the parent is identified.  The child mail should
        not have any subsequent log lines: only its parent should refer to
        it.

    \item If the mail is a child mail with an identified parent it is
        entered in the database as usual, then removed from its parent's
        list of children.  If this child is the last mail on that list, and
        the parent has already been entered in the database, the parent
        will be removed from the state tables.

    \item If the mail is a parent it is entered in the database as usual,
        because there will be no further log lines for it.  There may be
        child mails waiting to be entered in the database; these are
        entered as normal, and removed from the state tables.  If there are
        more child mails, the parent's removal from the state tables will
        be postponed until the last child has been entered.

@@

@@

The DATA phase of the \acronym{SMTP} conversation is where the headers and
body of the mail are transferred.  Sometimes there is a timeout or the
connection is lost during the DATA phase;\footnote{For brevity's sake
\textit{timeout\/} will be used throughout this section, but everything
applies equally to lost connections.} when this occurs Postfix will discard
the mail and the parser needs to discard the data associated with that
mail.  It seems more intuitive to save the mail's data to the database, but
if a timeout occurs there is no data available to save; the timeout is
recorded and saved with the connection instead.  To deal properly with
timeouts the TIMEOUT action does the following:

@@

    \item If one or more recipients have been accepted, Postfix will have
        allocated a queueid for the incoming mail, and there may be a mail
        in the state tables that needs to be discarded.  The timeout may
        have interrupted transfer of an accepted delivery attempt, or it
        may have occurred after a mail delivery attempt was rejected.  To
        distinguish between the two cases the action compares the timestamp
        of the last accepted mail against the timestamp of the last log
        line from \daemon{smtpd} for that connection; this depends on the
        CLONE action keeping a list of all mails accepted for each
        connection.  If the mail acceptance timestamp is later than the
        timeout timestamp, the timeout applies to the just-accepted mail,
        which will be removed from the state tables, without entering it in
        the database.  If the \daemon{smtpd} timestamp is later there was a
        rejection between the accepted mail and the timeout: the action
        assumes that the timeout applies to the rejected delivery attempt
        and finishes.  This assumption is not necessarily correct, because
        Postfix may have accepted an earlier recipient and rejected a later
        one, in which case the timeout applies to the accepted mail, which
        should be discarded; this has not been a problem in practice.
        Processing timeouts is further complicated by the presence of out
        of order \daemon{cleanup} log lines: see \sectionref{discarding
        cleanup log lines} for details.

@@

This complication is dealt with by the TIMEOUT action, with help from the
CLONE action.
